---
title: "Final"
author: "carter brendle"
date: "2025-12-11"
output: html_document
---

###Data Analytics Final - Drug resistance is an incredibly important aspect of our daily lifes interacting with modern medicine. This data set will be analyzing several parts of drug resistance including RBC, resistantdensity, and days infectious.

## Introduction

Drug resistance is a major challenge in treating infectious diseases. When pathogens evolve the ability to survive drug exposure, infections may last longer and transmission becomes more likely. Understanding how treatment influences resistance — and how resistance affects infectious duration — is crucial for public health and clinical strategy.

### Research Question

Does drug treatment reduce the evolution of resistance, and do drug-resistant infections last longer than non-resistant infections?

### Hypothesis

We expect:
1. Drug treatment will reduce resistant pathogen density.
2. Increased resistance will be result with longer infectious durations.

#Data and Methods - tidying and examing the drugresistance document
```{r setup, message = FALSE, warning = FALSE}
library(tidyverse)
library(mgcv)

drug <- read_csv("DrugResistance.csv")

glimpse(drug)
summary(drug)
```
```{r}
custom_colors <- c(
  "Aggressive" = "#F8766D",
  "Light" = "#7CAE00",
  "Moderate" = "#00BFC4",
  "Untreated" = "#C77CFF"
)

theme_custom <- theme_bw() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    legend.position = "none",
    panel.grid.minor = element_blank()
  )
```
```{r tidy-data}
drug_tidy <- drug %>%
  mutate(
    Treatment = factor(Treatment),                 # ensure categorical
    logResist = log10(ResistantDensity + 1),       # transform skewed outcome
    Infectious = DaysInfectious > 0                # create a logical variable
  )

glimpse(drug_tidy)
```
#So far the data has been slightly tidied and we've glimpsed it.
#Hypotheses and Testing
```{r aov-resist}
aov_resist <- aov(logResist ~ Treatment, data = drug_tidy)
summary(aov_resist)
aov_resist <- aov(logResist ~ Treatment, data = drug_tidy)
summary(aov_resist)
lm_mod <- lm(DaysInfectious ~ logResist + Treatment, data = drug_tidy)
summary(lm_mod)
```
#Above shows us R-Squared is 0.925 and p < 2.2r-16. Also, treatments differ in resistance levels, once resistance level is included, treatment no longer predicts infection duration. The original treatments effect isn't driving the duration, resistance evolution is.
#Resistant parasite density strongly predicted how long hosts remained infectious (β = 2.01, p < 0.0001). Hosts with higher resistance levels experienced substantially longer infections. After accounting for resistant density, treatment group was no longer a significant predictor of days infectious (all p > 0.10). This suggests that the primary driver of prolonged infection is the evolution of resistance, rather than the direct effect of the drug. These results indicate that drug treatment can unintentionally select for resistant strains that prolong infection and potentially increase disease transmission.

```{r check-names}
names(drug_tidy)
head(drug_tidy)
```
```{r prep-data}
drug_tidy <- drug_tidy %>%
  mutate(Treatment = factor(Treatment))
```


#Figure 1 - Log of ResistantDensity by Treatment Group.
```{r figure_logresist, echo=TRUE, message=FALSE, warning=FALSE, fig.width=7, fig.height=5}
ggplot(drug_tidy, aes(Treatment, logResist, fill = Treatment)) +
  geom_boxplot(alpha = 0.85) +
  scale_fill_manual(values = custom_colors) +
  labs(
    title = "Drug Treatment Reduces Resistant Parasite Density",
    x = "Treatment Group",
    y = "log10(Resistant Density + 1)"
  ) +
  theme_custom
```

#Figure 2 - Days Infectious by Treatment Group.
```{r fig_daysinfectious, echo=TRUE, message=FALSE, warning=FALSE, fig.width=7, fig.height=5}
ggplot(drug_tidy, aes(Treatment, DaysInfectious, fill = Treatment)) +
  geom_boxplot(alpha = 0.85) +
  scale_fill_manual(values = custom_colors) +
  labs(
    title = "Drug Treatment Shortens Infectious Period",
    x = "Treatment Group",
    y = "Days Infectious"
  ) +
  theme_custom
```

#Figure 3 - Days Infectious vs. Log ResistantDensity, color-coding is Treatment Group. From Untreated to Aggressive, colors go Purple, Green, Blue, and Red.
```{r fig_logresist_vs_days, echo=TRUE, message=FALSE, warning=FALSE, fig.width=7, fig.height=5}
ggplot(drug_tidy, aes(logResist, DaysInfectious, color = Treatment)) +
  geom_point(size = 3, alpha = 0.75) +
  geom_smooth(method = "lm", se = TRUE) +
  scale_color_manual(values = custom_colors) +
  labs(
    title = "Higher Resistance Density Leads to Longer Infection",
    x = "log10(Resistant Density + 1)",
    y = "Days Infectious"
  ) +
  theme_custom
```
#Interaction Model
```{r interaction_model}
lm_interact <- lm(DaysInfectious ~ logResist * Treatment,
                  data = drug_tidy)

summary(lm_interact)
```
#GAM Model
```{r gam_model}
gam_mod <- gam(DaysInfectious ~ s(logResist) + Treatment,
               data = drug_tidy)

summary(gam_mod)
```
```{r interact_plot, fig.width=6, fig.height=4}
# Prediction grid for interaction model
pred_int <- expand.grid(
  logResist  = seq(min(drug_tidy$logResist, na.rm = TRUE),
                   max(drug_tidy$logResist, na.rm = TRUE),
                   length.out = 100),
  Treatment  = levels(drug_tidy$Treatment)
)

pred_int$fit <- predict(lm_interact, newdata = pred_int)

ggplot(drug_tidy, aes(x = logResist, y = DaysInfectious, color = Treatment)) +
  geom_point(alpha = 0.3) +
  geom_line(data = pred_int,
            aes(y = fit),
            linewidth = 1.1) +
  labs(
    title = "Linear interaction model: resistance × treatment",
    x = "Log resistance",
    y = "Predicted days infectious"
  ) +
  theme_minimal()
```
```{r gam_plot_gg, fig.width=6, fig.height=4}
# Create a prediction grid over the range of logResist for each treatment
pred_grid <- expand.grid(
  logResist  = seq(min(drug_tidy$logResist, na.rm = TRUE),
                   max(drug_tidy$logResist, na.rm = TRUE),
                   length.out = 100),
  Treatment  = levels(drug_tidy$Treatment)
)

pred_grid$fit <- predict(gam_mod, newdata = pred_grid)

ggplot(drug_tidy, aes(x = logResist, y = DaysInfectious, color = Treatment)) +
  geom_point(alpha = 0.3) +
  geom_line(data = pred_grid,
            aes(y = fit),
            linewidth = 1.1) +
  labs(
    title = "GAM: Non-linear effect of resistance on infection duration",
    x = "Log resistance",
    y = "Predicted days infectious"
  ) +
  theme_minimal()
```
```{r}
#Figure X. The linear interaction model shows that infection duration increases with resistance, and that the intercepts (and possibly slopes) differ between treatment groups. This suggests that treatment not only shifts average infection length, but may also change how strongly resistance affects duration.
#Figure Y. The GAM fit reveals a non-linear relationship between resistance and infection duration: increases in log resistance have a stronger effect at lower values and then begin to level off. This supports the idea that simple straight-line models may miss important curvature in the biological response.
```

##Conclusion - Drug treatment was highly effective in reducing resistant parasite load and infection duration. Animals receiving aggressive and moderate treatments show improved outcomes. The severity of drug resistance was strongly predictive of infection duration. Higher resistant parasite density increases number of days infectious, suggest drug-resistant parasites are more taxing on host fitness and are critical for transmission. I guess my take away is while drug treatments do work, resistance undermines success by prolonging infectiousness.


