---
title: "exam 3"
author: "carter"
date: "2025-12-04"
output: html_document
---
#Chunk 1 - setup and browsing
```{r setup, message = FALSE}
library(tidyverse)

fac <- read.csv("FacultySalaries_1995.csv")
glimpse(fac)
head(fac)
```
#Chunk 2 - to tidy data and checking
```{r}
fac_tidy <- fac %>%
  pivot_longer(
    cols = c(AvgAssistProfSalary, AvgAssocProfSalary, AvgFullProfSalary),
    names_to = "Rank",
    values_to = "Salary"
  ) %>%
  mutate(
    Rank = recode(Rank,
                  "AvgAssistProfSalary" = "Assist",
                  "AvgAssocProfSalary"  = "Assoc",
                  "AvgFullProfSalary"   = "Full"),
    Rank = factor(Rank, levels = c("Assist", "Assoc", "Full"))
  )
glimpse(fac_tidy)
head(fac_tidy)
```
##Task 1 - a plot, showing salary plots by rank
```{r task1_plot, message = FALSE, warning = FALSE}
fac_tidy %>%
  ggplot(aes(x = Rank, y = Salary, fill = Rank)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.15, alpha = 0.3, size = 1) +
  facet_grid(~ Tier) +
  labs(x = "Rank", y = "Salary") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
##Task 2 - ANOVA using a one-way ANOVA with State, Tier, and Rank as independent variables and Salary as the response. It looks like all 3 are statistically significant on salary.
```{r anova_model, message = FALSE}
anova_mod <- aov(Salary ~ State + Tier + Rank, data = fac_tidy)
summary(anova_mod)
```
##Task 3 - Juniper Oils - I tidied the Juniper_Oils data by pivoting all chemical concentration columns into a long format, creating a compound column with the chemical name and a concentration column with its measured value.
```{r load_juniper}
juniper <- read_csv("Juniper_Oils.csv")
glimpse(juniper)
chem_cols <- c("alpha-pinene","para-cymene","alpha-terpineol","cedr-9-ene",
               "alpha-cedrene","beta-cedrene","cis-thujopsene",
               "alpha-himachalene","beta-chamigrene","cuparene","compound 1",
               "alpha-chamigrene","widdrol","cedrol","beta-acorenol",
               "alpha-acorenol","gamma-eudesmol","beta-eudesmol",
               "alpha-eudesmol","cedr-8-en-13-ol","cedr-8-en-15-ol",
               "compound 2","thujopsenal")
juniper_tidy <- juniper %>%
  pivot_longer(
    cols = all_of(chem_cols),
    names_to = "Compound",
    values_to = "Concentration"
  )
```
##Task 4 - I ran separate generalized linear models for each compound, testing whether concentration changes as YearsSinceBurn increases. I used broom::tidy() to extract coefficients and p-values. Chemicals with p < 0.05 for YearsSinceBurn were considered significantly affected by fire age.
```{r task4_plot, message = FALSE, warning = FALSE}
juniper_tidy %>%
  ggplot(aes(x = YearsSinceBurn, y = Concentration)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess", se = TRUE) +
  facet_wrap(~ Compound, scales = "free_y") +
  labs(x = "Years Since Burn", y = "Concentration") +
  theme_bw()

head(juniper_tidy)
```
##Task 5 - GLM. I fit separate generalized linear models for each chemical, using Concentration as the response and YearsSinceBurn as the predictor. I used `broom::tidy()` to extract coefficients and then filtered for the YearsSinceBurn term with p < 0.05. The resulting table shows which chemical concentrations are significantly affected by time since burn.
```{r task5_glm, message = FALSE, warning = FALSE}
library(broom)
library(dplyr)

# Get list of all compounds
compounds <- unique(juniper_tidy$Compound)

# Fit a glm for each compound and store results
res_list <- lapply(compounds, function(cmp) {
  dat <- juniper_tidy %>% filter(Compound == cmp)
  
  # fit GLM (gaussian family ~ same as linear model)
  m <- glm(Concentration ~ YearsSinceBurn, data = dat, family = gaussian())
  
  # tidy the model output and add the compound name as a column
  tidy(m) %>%
    mutate(Compound = cmp)
})

# Combine all results into one data frame
glm_results <- bind_rows(res_list)

# Keep only significant YearsSinceBurn terms
glm_sig <- glm_results %>%
  filter(term == "YearsSinceBurn",
         p.value < 0.05) %>%
  arrange(p.value)

glm_sig
```

